Query Optimization for Dynamic Imputation

### 2017-05-11 rebuttal.tex

  \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.
    \todo{}

  \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach — between 0 and 24 percent" 24 percent error is pretty bad.
    \todo{}

  \item S2: The drop operator appears to be a weaker form of SQL's null-value / 3-valued Boolean semantics. The latter might actually be more useful. For example, consider the case of a predicate (A $>$ 5 OR B $<$ 10). If A $>$ 5, then the condition is independent of B, and the tuple can still be salvaged. If A $\leq$ 5, then the expression evaluates to UNKNOWN and the tuple is dropped anyway. For aggregates there is a greater need, as one might have different aggregates computed over different subsets of the data, but even here it's not clear that this is any less informative than just throwing data away.

    \resp{We agree that it would be interesting to allow drops to be omitted before filters.
      This would increase the plan space somewhat and might lead to better plans in some cases.
      This optimization would only change the output of the filter when the predicate is disjunctive, so would not have an impact on the queries we used in our experiments.
      Filters with conjunctive predicates behave identically with and without the drop, although removing the drop will give a minor performance improvement.}

### 2017-05-11 rebuttal.tex

\documentclass{vldb}
\usepackage{xcolor}
\newcommand{\resp}[1]{{\color{blue}{#1}}}
\newcommand{\todo}[1]{{\color{red}{\textit{#1}}}}
\newcommand{\ProjectName}[0]{ImputeDB}

\begin{document}
\section{Multiple reviewers}
\begin{itemize}
\item Redundant figures, excessive whitespace, formatting.

  \resp{We've compressed many of the figures and removed whitespace to make room for additional material.}
\end{itemize}

  \section{Reviewer 1}
  \begin{itemize}
  \item W2. The paper completely obscures the ML technique being used: (1) It would be nice to have some discussion of the complete API for implementing an ML technique, as well as how one might implement a cost-estimator for one or two examples. (2) It would be helpful to see the impact of a few different approaches (e.g., Tree-based, Interpolation, Regression, FD-based) on performance and accuracy. (3) Given that the ML time estimator is being treated as a black box, it would be useful to see how sensitive the plan selector is to inaccurate estimates.

  \resp{
  We extended our evaluation to include two commonly used imputation methods: mean value imputation and hot-deck imputation. We also added
 details on how to pick the heuristic functions that define the API for impute operators. We provide details on how these functions are defined
 for all the methods used in our evaluation. 
  }

@@

  \item W3. Reducing runtime to ~10s is excellent, but comes at a very serious accuracy cost (3-5x worse SMAPE). Worse still, there's no way for the user to differentiate between the cases of Q4 and Q8 at runtime. 24\% error relative to direct imputation is huge.

    \resp{
      The outsized error in Q8 is due to a drop operator placed very early on in the plan (a result of optimizing for performance). This produced significantly smaller sample sizes for various
      groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
      using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.
    }

  \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.
   \resp{
   Yang's work is definitely related, and we've added it to our related work section. However, it is worth noting that their decision process for application of lenses is at the tuple-level,
   rather than with respect to the query plan.
   }

  \item S1: "Furthermore, the results obtained with this approach are similar in quality to those obtained with the traditional imputation approach between 0 and 24 percent" 24 percent error is pretty bad.

@@

    \resp{We have clarified this in the paper by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to \textit{Penalty}.}

  \item S3.4.3: An extremely nice emergent property from the L-value function for the imputation operator seems to be that the operator will be moved to a position in the plan where it has access to the maximum training data. I am slightly concerned that by focusing entirely on size, this might lead to skew in the training set. For example, consider a one-to-many join \verb/µ(R) |X| S/. Your optimization algorithm would likely wish to move imputation up: \verb/µ(R |X| S)/. Yes, there is more training data, but if the fanout for each tuple is skewed, some tuples in R will become over-represented in the join result. A similar observation holds for selection. For example take your Q8, where the conditions "studentdebtowe $>$ 0 AND schooldegree $\geq$ 0" presumably indicate invalid data that should not be used for training. (is this perhaps why you have so much error for full imputation on Q8?)

    \resp{This is a reasonable concern. Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query requires the construction a large, skewed data set. The problem seems to be particular to queries which generate large intermediate tables. In some cases, as when joining two datasets, the intermediate table is a good target for training because it exposes correlations which would be hidden in the separate tables. In other cases, this intermediate table is large and skewed, which causes problems for imputation. Avoiding this case requires us to estimate the skew in these intermediate tables.

     To clarify the point about Q8, we've removed the predicate condition $schooldegree \geq 0$. 
      We originally enumerated strings with positive integers and replaced missing values strings with 0, rather than perform imputation on them, so every enumerated value
      satisfied this predicate trivially. We have changed our evaluation to impute string enumerations as well and dropped this predicate.
      }
    %\todo{Need a better explanation here.}

@@

    \resp{
      We have added two new imputation methods to the experiments, to show that the planning algorithm can be used successfully with different imputation methods.
      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving rules for choosing cost functions for a new imputation method.
    }
  \end{itemize}

  \section{Reviewer 3}
  \begin{itemize}
  \item W2. More work is required in the presentation: important parts are not explained in depth, while there are many details in other sections that can be omitted.

  \item W3. More experiments needed.

  \resp{
  We've extended the evaluation section to include more impute operators and an evaluation of planning times.
  }

@@

  \item D1. Section 3, which is the meat of the paper, is very overloaded, without getting in depth in various parts of the system. It definitely should be split in more sections, and more details should be given. In particular:
    -- Give some more information about cardinality estimation. Also, Algorithm 3 is too far from the text, and needs some explanation too.
    -- Some more details are also required in the Imputation quality section. Why are you choosing these specific formulas for Loss, especially for the impute operator. Some examples would help.
    -- Algorithms 1 and 2 could probably be presented in an easier to grasp way.

    \resp{
    We've broken up this section and consolidated algoirthm notation. The cardinality estimation section provides more detail about the algorithm. We also
    added intuition for how the heuristic Penalty (a renaming of Loss, for clarity) and Time should be designed. We walk through these definitions for three different
    impute operators.
    }

### 2017-05-11 paper/algorithm.tex

\subsubsection{Cardinality estimation}
\label{sec:cardinal}
The computation of $\textsc{Penalty}(q)$ and$\textsc{Time}(q)$ relies on accurate cardinality estimates for the plan $q$ and its sub-plans.
These estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.

To compute cardinality, we maintain histograms for each column in the database.
During query planning, each of the logical nodes in a query plan points to a set of histograms which describe the distribution of values in the output of the node.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.

\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.
Drop operators always reduce the cardinality of the relation. Given $\delta_C$, each column $c \in C$ clears its count of tuples with missing values,
which can yield distinct cardinality estimates across attributes. To provide consistent estimates, we take the lowest such estimate $e$ and
scale all attribute histograms based on their distribution such that each provides an estimate of $e$. In the case of $\mu_C$, the cardinality estimate
remains constant. The count of missing values across attributes is cleared and redistributed along the appropriate histograms based on their distribution.
In the case of $\sigma$ and $\bowtie$ operations, we estimate the cardinality of the output relation and adjust histogram counts based on their distribution to sum to that estimate.

\begin{algorithm}
\input{algorithms/histogram-transform.tex}
\end{algorithm}

@@

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$, where $T$ is a set of tables and $p$ is a query plan}
  \label{fig:semantics-plan-cache}
\end{figure}

>>>>

\subsubsection{Cardinality estimation}
\label{sec:cardinal}
The computation of $\textsc{Penalty}(q)$ and $\textsc{Time}(q)$ relies on accurate cardinality estimates for the plan $q$ and its sub-plans.
These estimates are impacted not just by filtering or joining, as in the traditional relational algebra, but also by the imputation operators.

To compute cardinality, we maintain histograms for each column in the database.
These histograms track the distribution of values in the column as well as the number of missing values.
During query planning, each of the logical nodes in a query plan points to a set of histograms which describe the distribution of values in the output of the node.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as described in~\Cref{algo:histogram-transformation} to account for the new operation in the plan.

Drop operators (\Crefrange{lst:drop-start}{lst:drop-end}) reduce the cardinality of their input relation by removing tuples which contain missing values.
For a drop $\delta_C$, for each column $c \in C$ we set its missing value count to zero, because all of these tuples will be dropped.
This can result in a set of histograms which have different cardinalities if some columns have more missing values than others.
To account for the discrepancy, we rescale all the histograms without changing the shape of their distributions to match the lowest cardinality column.

Impute operators (\Crefrange{lst:impute-start}{lst:impute-end}) maintain the cardinality of their inputs, but they also fill in missing values.
For an impute $\mu_C$, we set the missing value count to zero, and rescale each histogram so that its new cardinality, with no missing values, is the same as its old cardinality.
We assume that the imputed values will not change the shape of the distribution of values in each column.

Cardinality estimation for standard relational operators (\Crefrange{lst:std-start}{lst:std-end}) is the same as in a normal database.
\begin{algorithm}
\input{algorithms/histogram-transform.tex}
\end{algorithm}

@@

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
    \text{ where }\ P' &= \left\{q ~\Bigg|~ \begin{aligned}
        q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
        \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
        \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
      \end{aligned}
    \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$, where $T$ is a set of tables and $p$ is a query plan}
  \label{fig:semantics-plan-cache}
\end{figure}

### 2017-05-11 paper/relatedwork.tex

>>>>

\todo{confirm that this reading of Yang is correct}
\cite{yang2015lenses} develops an incremental approach to ETL data cleaning, through the use
of probabilistic compositional repair operators, called Lenses. These can be used
to repair missing values by imposing data constraints such as \verb|NOT NULL|. 
Similarly to \ProjectName{}, they provide a way to balance between the quality of the cleaning 
performed (expressed by the confidence of their output) and the cost of those operations. However, their 
repairs take place on a per-tuple basis, with the non-deterministic layer of their system replacing place-holder
variables with inferred values. Meanwhile, \ProjectName{} optimizes imputation at the plan-level and
provides the same impute technique for all tuples streaming through a given operator.
